---
title: "Standardabweichung berechnen"
format: 
  html:
    toc: true
    number-sections: true
    code-tools: true
  pdf:
    toc: true
---

Standard deviation ($\sigma$) equals the squareroot ($\sqrt{}$) of the squared ($^2$) sum ($\sum$) of the difference between each value and the mean ($x_{1...n}-\mu$), divided by the number of values minus 1 ($N-1$):

\begin{align}
\sigma = \sqrt{\frac{\sum(x_i - \mu)^2}{N-1}}
\end{align}

Let's try doing this step-by-step:

1. Save values as object
1. Determine $N$
2. compute mean ($\mu$)
3. compute difference between each value and the mean (subtract mean from each value; $x_{1...n}-\mu$)
4. square these values ($^2$)
5. sum the squared values ($\sum$)
5. divide this value by N-1
6. compute the squareroot ($\sqrt{}$)

# Save values as object

```{r}
werte <- c(13, 0, 2000, 5, 19)
```

Print values

```{r}
werte
```

# Determine $N$

#### By hand: print values and count how many there are  {.unnumbered}
```{r}
werte
```

#### Or, use the `length()` function  {.unnumbered}
```{r}
length(werte)
```


# Compute mean

The mean ($\mu$) is the sum ($\sum$) of the values ($x$) divided by the number of values ($N$):

$$
\mu = \frac{\sum x}{N}
$$

#### By hand {.unnumbered}

Typing out all the values:

```{r}
(13 + 0 + 2000 + 5 + 19)/5
```

#### Or by using the functions `sum()` and `length()`  {.unnumbered}
```{r}
sum(werte)/length(werte)
```

#### Or by simply using the function `mean()`  {.unnumbered}

```{r}
mean(werte)
```

# Subtract mean from each value

This can also be done several different ways

#### By hand (tedious and prone to human error)  {.unnumbered}
```{r}
(13 - 407.4)
(0 - 407.4) 
(2000 - 407.4)
(5 - 407.4)
(19 - 407.4)
```

#### Or using the object name  {.unnumbered}
```{r}
werte - mean(werte)
```

# Square these values

#### By hand  {.unnumbered}
```{r}
(13 - 407.4)^2
(0 - 407.4)^2
(2000 - 407.4)^2
(5 - 407.4)^2
(19 - 407.4)^2
```

#### Or by using `^2` with the variable name  {.unnumbered}
```{r}
(werte - mean(werte))^2
```

# Sum the squared values

#### By hand  {.unnumbered}
```{r}
(13 - 407.4)^2 +
(0 - 407.4)^2 +
(2000 - 407.4)^2 +
(5 - 407.4)^2 +
(19 - 407.4)^2
```

#### Or with the object name and some functions  {.unnumbered}
```{r}
sum((werte - mean(werte)) ^ 2)
```

# Divide by N-1

#### By hand  {.unnumbered}

```{r}
((13 - 407.4) ^ 2 +
   (0 - 407.4) ^ 2 +
   (2000 - 407.4) ^ 2 +
   (5 - 407.4) ^ 2 +
   (19 - 407.4) ^ 2)/(5-1)
```

#### Or with the object name and some functions  {.unnumbered}

```{r}
sum((werte - mean(werte)) ^ 2) / (length(werte) - 1)
```

# Compute the squareroot

#### By hand  {.unnumbered}

```{r}
sqrt(((13 - 407.4)^2 +
(0 - 407.4)^2 +
(2000 - 407.4)^2 +
(5 - 407.4)^2 +
(19 - 407.4)^2)/(5-1))
```

#### Or with the object name and some functions  {.unnumbered}

```{r}
sqrt(sum((werte - mean(werte)) ^ 2) / (length(werte) - 1))
```

# Check with `sd()` function

```{r}
sd(werte)
```

