---
title: "Quick review"
format:
  html:
    output-file: wiederholung_blatt.html
    number-sections: true
    number-depth: 3
    toc: true
    toc-title: "heutige Themen"
    code-overflow: wrap
    code-tools: true
    self-contained: true
  revealjs: 
    output-file: wiederholung_folien.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    number-depth: 1
    toc: true
    toc-depth: 1
    toc-title: 'Überblick'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    # chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  pdf:
    output-file: wiederholung1.pdf
    toc: true
    number-sections: true
    colorlinks: true
    code-overflow: wrap
---

# Wiederholung

- loading packages
- loading data
- wrangling
- code style

## Pakete

```{r}
pacman::p_load(
  tidyverse,
  here
)
```

## Einlesung

To read in data we need to:
  - use the `read_csv()` package from `readr` ( a `tidyverse` package)
  - use `here::here()` to tell R to look in our project folder
  - list the folder our data is in: `daten`
  - give the *exact name* of the file stored in `daten`: e.g., `flights.csv`
  
```{r}
#| eval: true
#| echo: true
#| results: hide
#| code-line-numbers: "|1|"
df_flights <-
  read_csv(
    here::here(
      "daten",
      "flights.csv"
    )
  )
```

## Wrangling

Why does the following return no observations?

```{r}
#| eval: false
#| echo: true
df_flights %>%
  filter(arr_delay > 120,
         dep_delay < 0,
        dest=="HOU" | 
        dest=="IAH",
        carrier=="UA" |
        carrier=="AA" |
        carrier=="DL",
        month== 7 | 8 | 9
         )
# such flights don't exist
```

## Coding style

This code is a bit hard to read.

```{r}
#| eval: false
#| echo: true
df_flights %>% filter(arr_delay > 120, dep_delay == 0, dest == "IAH" | dest == "HOU", carrier == "UA" | carrier == "AA" | carrier == "DL", month == 07 | month == 08 | month == 09)
```

---

This code is easier to read, although it takes up more space. This is generally how code *should* be structured.

```{r}
#| eval: false
#| echo: true
df_flights %>%
  filter(
    arr_delay > 120,
    dep_delay == 0,
    dest == "IAH" | dest == "HOU",
    carrier == "UA" |
      carrier == "AA" |
      carrier == "DL",
    month == 07 |
      month == 08 |
      month == 09
  )
```

---

A simple way to take 'messy' code is to highlight it, and use the shortcut `Cmd/Strg+Shift+A`

